---
output:
  html_document: default
  pdf_document: default
---
White Wine Quality Exploration by Akshat N. Tickoo
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(knitr)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine_data = read.csv('wineQualityWhites.csv')
```

This report explores a dataset containing data about different components of wine taken from 4898 wine samples.

# Univariate Plots Section
```{r echo = FALSE}
str(wine_data)
summary(wine_data)
```

The dataset consists of 4989 observations with 12 variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
ggplot(aes(x = fixed.acidity), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$fixed.acidity), 
                                max(wine_data$fixed.acidity)), 
                     breaks = seq(min(wine_data$fixed.acidity), 
                                  max(wine_data$fixed.acidity), 1))
summary(wine_data$fixed.acidity)
quantile(wine_data$fixed.acidity, 0.95)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$volatile.acidity), 
                                max(wine_data$volatile.acidity)), 
                     breaks = seq(min(wine_data$volatile.acidity), 
                                  max(wine_data$volatile.acidity), 0.1))
summary(wine_data$volatile.acidity)
quantile(wine_data$volatile.acidity, 0.95)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$citric.acid), 
                                max(wine_data$citric.acid)), 
                     breaks = seq(min(wine_data$citric.acid), 
                                  max(wine_data$citric.acid), 0.1))
summary(wine_data$citric.acid)
quantile(wine_data$citric.acid, 0.95)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = chlorides), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$chlorides), 
                                max(wine_data$chlorides)), 
                     breaks = seq(min(wine_data$chlorides), 
                                  max(wine_data$chlorides), 0.05))
summary(wine_data$chlorides)
quantile(wine_data$chlorides, 0.95)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = free.sulfur.dioxide), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$free.sulfur.dioxide), 
                                max(wine_data$free.sulfur.dioxide)), 
                     breaks = seq(min(wine_data$free.sulfur.dioxide), 
                                  max(wine_data$free.sulfur.dioxide), 15))
summary(wine_data$free.sulfur.dioxide)
quantile(wine_data$free.sulfur.dioxide, 0.95)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = total.sulfur.dioxide), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$total.sulfur.dioxide), 
                                max(wine_data$total.sulfur.dioxide)), 
                     breaks = seq(min(wine_data$total.sulfur.dioxide), 
                                  max(wine_data$total.sulfur.dioxide), 50))
summary(wine_data$total.sulfur.dioxide)
quantile(wine_data$total.sulfur.dioxide, 0.95)
```


Clearly these 6 variables have outliers as most of them (95%) are within a suitable range. If the outliers are removed we get normal distributions instead of skewed ones as described by the following.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(gridExtra)

p1 = ggplot(aes(x = fixed.acidity), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$fixed.acidity), 
                                quantile(wine_data$fixed.acidity, 0.95)), 
                     breaks = seq(min(wine_data$fixed.acidity), 
                                  quantile(wine_data$fixed.acidity, 0.95), 0.5))

p2 = ggplot(aes(x = volatile.acidity), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$volatile.acidity), 
                                quantile(wine_data$volatile.acidity, 0.95)), 
                     breaks = seq(min(wine_data$volatile.acidity), 
                                  quantile(wine_data$volatile.acidity, 0.95), 0.1))

p3 = ggplot(aes(x = citric.acid), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$citric.acid), 
                                quantile(wine_data$citric.acid, 0.95)), 
                     breaks = seq(min(wine_data$citric.acid), 
                                  quantile(wine_data$citric.acid, 0.95), 0.1))

p4 = ggplot(aes(x = chlorides), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$chlorides), 
                                quantile(wine_data$chlorides, 0.95)), 
                     breaks = seq(min(wine_data$chlorides), 
                                  quantile(wine_data$chlorides, 0.95), 0.01))

p5 = ggplot(aes(x = free.sulfur.dioxide), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$free.sulfur.dioxide), 
                                quantile(wine_data$free.sulfur.dioxide, 0.95)), 
                     breaks = seq(min(wine_data$free.sulfur.dioxide), 
                                  quantile(wine_data$free.sulfur.dioxide, 0.95), 5))

p6 = ggplot(aes(x = total.sulfur.dioxide), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$total.sulfur.dioxide), 
                                quantile(wine_data$total.sulfur.dioxide, 0.99)), 
                     breaks = seq(min(wine_data$total.sulfur.dioxide), 
                                  quantile(wine_data$total.sulfur.dioxide, 0.99), 50))

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 3)
```

While these four are different in their orientation. Residual Sugar, Sulphates and Alcohol levels are positively skewed while there are some outliers which are quite high in the graph of Density.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$residual.sugar), 
                                quantile(wine_data$residual.sugar, 0.95)), 
                     breaks = seq(min(wine_data$residual.sugar), 
                                  quantile(wine_data$residual.sugar, 0.95), 2))

ggplot(aes(x = sulphates), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$sulphates), 
                                max(wine_data$sulphates)), 
                     breaks = seq(min(wine_data$sulphates), 
                                  max(wine_data$sulphates), 0.2))

ggplot(aes(x = alcohol), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$alcohol), 
                                max(wine_data$alcohol)), 
                     breaks = seq(min(wine_data$alcohol), 
                                  max(wine_data$alcohol), 1))

ggplot(aes(x = density), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$density), 
                                quantile(wine_data$density, 0.95)), 
                     breaks = seq(min(wine_data$density), 
                                  quantile(wine_data$density, 0.95), 
                                  0.001))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = pH), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$pH), 
                                max(wine_data$pH)), 
                     breaks = seq(min(wine_data$pH), 
                                  max(wine_data$pH), 0.2))
summary(wine_data$pH)
quantile(wine_data$pH, 0.95)
```

While the graph of pH seems like a clear normal distribution without any outliers as denoted by the summary as well.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality), data = wine_data) + 
  geom_bar() + 
  scale_x_continuous(limits = c(min(wine_data$quality), 
                                max(wine_data$quality)), 
                     breaks = seq(min(wine_data$quality), 
                                  max(wine_data$quality), 1))

wine_data$quality_level <- ifelse(wine_data$quality <= 4, 'bad', 
                                  ifelse(wine_data$quality <= 7, 'medium', 
                                         'good'))
```

Quite evidently, most of the samples have a quality between 5-7. I've made 3 quality levels of the wine depending on their quality score, grouping them into 'good', 'medium' and 'bad'.

# Univariate Analysis

### What is the structure of your dataset?
There are 4898 wine samples in this dataset having 12 variables (Fixed Acidity, Volatile Acidity, Citric Acid, Residual Sugar, Chlorides, Free Sulphur Dioxide, Total Sulphur Dioxide, Density, pH level, Sulphates, Alcohol and Quality) all numerical, with most of them being quantities. 

### What is/are the main feature(s) of interest in your dataset?
The main features seem to be the acidic levels, the sulphate levels and the alcohol levels.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
The free sulphur dioxide and the total sulphur dioxide can also be of support in further investigation of this dataset.

### Did you create any new variables from existing variables in the dataset?
Yes, I've made a variable quality_level, which groups the wines into 3 different types - 'good', 'medium' and 'bad', according to their quality scores.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
Yes, a lot many graphs seemed to be positively skewed when first plotted, but zooming in, we saw that they just have some outliers. Trimming the axis accordingly, and taking into consideration 95% of the values, the graphs on being plotted again seemed to have a normal distribution. 


# Bivariate Plots Section
Let's look at the correlations of all the variables with each other and how strong they are.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots, fig.width=15, fig.height=15}
library(GGally)
theme_set(theme_minimal(20))

set.seed(2016)
ggpairs(wine_data[sample.int(nrow(wine_data), 1000), ])
```

As seen in the above graph, we have the following considerable correlations - 
1. Negative 0.469 between pH and fixed acidity.
2. Residual sugar shows correlation with total sulphur dioxide (0.419), a strong correlation with density (0.831) and a negative correlation with alcohol (-0.467).
3. Total sulphur dioxide also shows strong correlations with density (0.583) and alcohol (-0.505).
4. Alcohol then again shows a strong negative correlation with density (-0.804).

So now we'll have to see how these main features affect the quality scores of the wine samples, one feature at a time!

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=15, fig.height=20}
p1 = ggplot(aes(pH), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

p2 = ggplot(aes(alcohol), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

p3 = ggplot(aes(fixed.acidity), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

p4 = ggplot(aes(total.sulfur.dioxide), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

p5 = ggplot(aes(free.sulfur.dioxide), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

p6 = ggplot(aes(density), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

p7 = ggplot(aes(residual.sugar), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

p8 = ggplot(aes(chlorides), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

p9 = ggplot(aes(volatile.acidity), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

p10 = ggplot(aes(citric.acid), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

p11 = ggplot(aes(sulphates), data = wine_data) + 
  geom_density(alpha = 2/3, aes(fill = quality_level))

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, ncol = 2)
```

Let's look at some of these closely using boxplots

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 = ggplot(aes(x = quality_level, y = fixed.acidity), data = wine_data) + 
  geom_boxplot()

p2 = ggplot(aes(x = quality_level, y = pH), data = wine_data) + 
  geom_boxplot()

grid.arrange(p1, p2, ncol = 2)
```

A very loose opposite relation between the two can be seen as confirmed earlier by the correlation coefficient of -0.469.

This also suggests that better quality wines tend to have a slightly higher pH level.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=15, fig.height=15}
p1 = ggplot(aes(x = quality_level, y = residual.sugar), data = wine_data) + 
  geom_boxplot()

p2 = ggplot(aes(x = quality_level, y = total.sulfur.dioxide), data = wine_data) + 
  geom_boxplot()

p3 = ggplot(aes(x = quality_level, y = residual.sugar), data = wine_data) + 
  geom_boxplot()

p4 = ggplot(aes(x = quality_level, y = density), data = wine_data) + 
  geom_boxplot()

p5 = ggplot(aes(x = quality_level, y = residual.sugar), data = wine_data) + 
  geom_boxplot()

p6 = ggplot(aes(x = quality_level, y = alcohol), data = wine_data) + 
  geom_boxplot()

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2)
```

Not much of relationship is visible here except a strong negative correlation with alcohol as seen in the last graph.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 = ggplot(aes(x = quality_level, y = total.sulfur.dioxide), data = wine_data) + 
  geom_boxplot()

p2 = ggplot(aes(x = quality_level, y = density), data = wine_data) + 
  geom_boxplot()

grid.arrange(p1, p2, ncol = 2)
```

Again not much of a visible correlation as suggested by our calculated coefficient of 0.583.

Let's look at the relation between alcohol and density, as all our graphs yet suggest a strong negative correlation between the two.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 = ggplot(aes(x = quality_level, y = alcohol), data = wine_data) + 
  geom_boxplot()

p2 = ggplot(aes(x = quality_level, y = density), data = wine_data) + 
  geom_boxplot()

grid.arrange(p1, p2, ncol = 2)
```

There we have it! As suggested by our calculated score of -0.804. 

This graph also suggests that the higher the alcohol content of the wine, the better its quality and vice versa. Similarly, the lower the density of water and sugar content in the wine, the better it tastes.

Let's look at two other features, whose density plots earlier intrigued me!

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=10}
p1 = ggplot(aes(x = quality_level, y = chlorides), data = wine_data) + 
  geom_boxplot()

p2 = ggplot(aes(x = quality_level, y = free.sulfur.dioxide), data = wine_data) + 
  geom_boxplot()

grid.arrange(p1, p2, nrow = 2)
```

Better quality wines tend to have a slightly lower chloride level. Similarly they tend to have slightly more free sulphur dioxide content in them.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
The most important relationship which came into light in this section was that between alcohol content and the quality of wine. Also density and the pH value also does seem to have an effect on the quality.

Rest of the features in the dataset have mostly flat graphs as seen above.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
Yes as we saw in the last, chlorides and free sulphur dioxide seem to also have minor effects on the quality.

### What was the strongest relationship you found?
The strongest relation is definitely that between the alcohol content and the quality of wine.



# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
